F\# Data: \textnormal{Making structured data first-class citizens}

### 2015-02-27 icfp-2015/paper.tex
Spellechecking

The XML format is similar to JSON in that it uses nested elements rather than a flat CSV-like 
structure. In our inference algorithm, we use the notion of \emph{records} for both JSON records 
and XML elements. However, unlike in JSON, the XML elements also have a name. An element can 
contain a collection of child elements. In case of XML, the collection can often be heterogeneous 
containing child nodes of different names. Consider the following (simplified) RSS news feed as an example:
%
{\small{
\begin{verbatim}
  <rss version="2.0"><channel>
    <title> BBC News - Europe </title>
    <item><title> Kurdish activists 
        killed in Paris </title></item>
    <item><title> German MPs warn 
        over UK EU exit </title></item>
  </channel></rss>
\end{verbatim}
}}
%
\noindent
The \ident{channel} node contains a single \ident{title} node (with the name of the feed) and multiple 
\ident{item} nodes (individual news articles). In RSS, the \ident{title} nodes always contain plain text,
while \ident{item} nodes contain \ident{title} (and also \ident{url} and othe properties omitted here).

@@

The grammar below defines \emph{structural type} $\sigma$. We distinguish between \emph{non-nullale types}
that always have a valid value (written as $\hat{\sigma}$) and \emph{nullable types} that encompass missing 
and \kvd{null} values (written as $\sigma$). We write $\nu$ for record field names and $\nu_{\ident{opt}}$
for record names, which can be empty:
%
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu_{\ident{opt}}\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{decimal} \lsep \ident{int} \lsep \ident{bit} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \hat{\sigma}\;\,\kvd{option} \lsep [\sigma] \\[0.1em]
              &\narrow{|}& \sigma_1 + \ldots + \sigma_n \lsep ~\top~ \lsep \kvd{null}
\end{array}
\end{equation*}
%
The non-nullable types include records (consisting of an optional name and zero or more fields with
their types) and primitive types. Records arising from XML documents are always named, while records
used by JSON and CSV providers are always unnamed.

@@

Finally, a union type in our model implicitly permits the \kvd{null} value. This is because structural union 
types are \emph{not} mapped to F\# union types. Instead our encoding requires the user to handle the 
cases one-by-one and so they also need to handle the situation when none of the cases matches (as discussed
in Section~\ref{sec:impl-unions}, this is the right choice for an F\# type provider, but it would not 
necessarilly be a good fit for other languages).

@@

\noindent
When finding a common supertype of two records (\emph{record-1}), we return a record type that has the union
of fields of the two arguments. We assume that the names of the first $k$ fields are the same and the
remaining fields have different names. To get a record with the right order of elements, we provide the
(\emph{order}) rule. The types of shared fields become common supertypes of their 
respective types (recursively). Fields that are present in only one record are marked as optional using
the following helper definition:
%
\begin{equation*}
\begin{array}{rcll}
 \addopt{\hat{\sigma}} &\narrow{=}& \hat{\sigma}~\kvd{option} \qquad&(\textnormal{non-nullable types})\\
 \addopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
\end{array}
\end{equation*}
%
Finding a common supertype of when one type is a union is more complicated. Our definiion aims to 
use a flat structure (avoid nesting unions) and limit the number of cases. This is done by grouping
types that have a common supertype which is not a union type. For example, rather than inferring 
$\ident{int} + (\ident{bool} + \ident{decimal})$, our algorithm will find the common supertype of
$\ident{int}$ and $\ident{decimal}$ (which is \ident{decimal}) and it will produce just $\ident{decimal}+\ident{bool}$.

@@

\item \textbf{Conversions.} Additional conversions are not strictly required, but prove useful in practice.
  These include converting to a smaller numerical type (such as float to integer) written as $i=\lfloor f\rfloor$,
  which is not defined in case of arithmetic overflow; conversion to and from string values written as
  ``$t$~represents~$i$'', which parses or formats a non-string value and also trating of decimals 
  0.0 and 1.0 as Booleans.

@@

\paragraph{Example 2.} The second example illustrates the remaining parts of the type provision,
including collections and union types. Teusing the \ident{Person} type from the previous example,
consider a collection $[\ident{Person}+\ident{string}]$, which contains a mix of \ident{Person} and string values.

@@

\subsection{Infering types from values}
\label{sec:formal-inferval}

@@

The key question is, what is a representative sample? Given a set of sample documents, 
the provided type is guaranteed to work if the inferred type of the input is a subtype of any of the
sample documents. Going back to Section~\ref{sec:inference-subtyping}, this means that:
%
\begin{itemize}[noitemsep]
\item[--] Input can contain smaller numerical values (for example, if a sample contains float, the input can contain an integer).
\item[--] Records in the actual input can have additional fields
\item[--] Records in the actual input can have fewer fileds, provided that the type of the fields is marked as optional in the sample
\item[--] Union types in the input can have both fewer or more cases
\item[--] When we have a union type in the sample, the actual input can also contain just values of one of the union cases
\end{itemize}
%
The following lemma states that the provided code (generated in Figure~\ref{fig:tp-generation})
works correctly on inputs $s'$ that is a subtype of the sample $s$. More formally, we require that the 
provided expression (using $s'$ as the input) can be reduced to a value and, if it is a class,
all its members can also be reduced to values.

@@

\noindent
Now that we know that the provided types are correct with respect to the subtyping relation, we can
look at the main theorem of the paper. It statest that, for any input (which is a subtype of any of the
samples) and any expression $e$, a well-typed program that uses the provided types does not ``go wrong''.

@@

The Lemma~\ref{thm:tp-correctness} guarantees that this does not happen. In Theorem~\ref{thm:safety},
we carefuly state that we only consider expressions $e_c$ which ``[do] not contain any structural 
values as sub-expressions''. This makes sure that the only code working with structural values is the
code generated by the type provider.

@@

The \emph{relativized safety} property does not guarantee the same amount of safety as standard type safety
for programming langauges without type providers. However, it reflects the reality of programming with external
data sources that is increasingly important in the age of web \cite{age-of-web}. So, type providers do not
reduce the safety -- they simply make the existing issues visible.

@@

Finally, the formal model presented here ignores two aspects of the real type provider mechanisms. As discussed
here, we do not use the \emph{type erasure} mechanism, but treat type providers as if they were actually 
generating code. For XML, CSV and JSON, both models can be used. However, one interesting aspect of erasing 
type providers is that the types can be generated lazilly -- only classes that are actually used in the type
checked code are generated. Extending our formalism to capture this aspect is an interesting future work
relevant to other type providers -- including the World Bank type provider that is also available
in the F\# Data library.

@@

\subsection{Parsing strucutred values}
\label{sec:impl-parsing}

@@

The type provider mechanism has been described before, but this paper is novel in that it explores concrete
type providers from the perspective of programming language theory. This is particularly important as the
F\# Data library is becoming de-facto standard tool for data access in F\#\footnote{At the time of writing, 
the library has 37,000 downloads on NuGet; over 1600 commits and 36 contributors on GitHub.} and other 
languages are begining to adopt mechanisms similar to F\# type providers.

>>>>

The XML format is similar to JSON in that it uses nested elements rather than a flat CSV-like 
structure. In our inference algorithm, we use the notion of \emph{records} for both JSON records 
and XML elements. However, unlike in JSON, the XML elements also have a name. An element can 
contain a collection of child elements. In case of XML, the collection can often be heterogeneous 
containing child nodes of different names. Consider the following (simplified) RSS news feed as an example:
%
{\small{
\begin{verbatim}
  <rss version="2.0"><channel>
    <title> BBC News - Europe </title>
    <item><title> Kurdish activists 
        killed in Paris </title></item>
    <item><title> German MPs warn 
        over UK EU exit </title></item>
  </channel></rss>
\end{verbatim}
}}
%
\noindent
The \ident{channel} node contains a single \ident{title} node (with the name of the feed) and multiple 
\ident{item} nodes (individual news articles). In RSS, the \ident{title} nodes always contain plain text,
while \ident{item} nodes contain \ident{title} (and also \ident{url} and other properties omitted here).

@@

The grammar below defines \emph{structural type} $\sigma$. We distinguish between \emph{non-nullable types}
that always have a valid value (written as $\hat{\sigma}$) and \emph{nullable types} that encompass missing 
and \kvd{null} values (written as $\sigma$). We write $\nu$ for record field names and $\nu_{\ident{opt}}$
for record names, which can be empty:
%
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu_{\ident{opt}}\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{decimal} \lsep \ident{int} \lsep \ident{bit} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \hat{\sigma}\;\,\kvd{option} \lsep [\sigma] \\[0.1em]
              &\narrow{|}& \sigma_1 + \ldots + \sigma_n \lsep ~\top~ \lsep \kvd{null}
\end{array}
\end{equation*}
%
The non-nullable types include records (consisting of an optional name and zero or more fields with
their types) and primitive types. Records arising from XML documents are always named, while records
used by JSON and CSV providers are always unnamed.

@@

Finally, a union type in our model implicitly permits the \kvd{null} value. This is because structural union 
types are \emph{not} mapped to F\# union types. Instead our encoding requires the user to handle the 
cases one-by-one and so they also need to handle the situation when none of the cases matches (as discussed
in Section~\ref{sec:impl-unions}, this is the right choice for an F\# type provider, but it would not 
necessarily be a good fit for other languages).

@@

\noindent
When finding a common supertype of two records (\emph{record-1}), we return a record type that has the union
of fields of the two arguments. We assume that the names of the first $k$ fields are the same and the
remaining fields have different names. To get a record with the right order of elements, we provide the
(\emph{order}) rule. The types of shared fields become common supertypes of their 
respective types (recursively). Fields that are present in only one record are marked as optional using
the following helper definition:
%
\begin{equation*}
\begin{array}{rcll}
 \addopt{\hat{\sigma}} &\narrow{=}& \hat{\sigma}~\kvd{option} \qquad&(\textnormal{non-nullable types})\\
 \addopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
\end{array}
\end{equation*}
%
Finding a common supertype of when one type is a union is more complicated. Our definition aims to 
use a flat structure (avoid nesting unions) and limit the number of cases. This is done by grouping
types that have a common supertype which is not a union type. For example, rather than inferring 
$\ident{int} + (\ident{bool} + \ident{decimal})$, our algorithm will find the common supertype of
$\ident{int}$ and $\ident{decimal}$ (which is \ident{decimal}) and it will produce just $\ident{decimal}+\ident{bool}$.

@@

\item \textbf{Conversions.} Additional conversions are not strictly required, but prove useful in practice.
  These include converting to a smaller numerical type (such as float to integer) written as $i=\lfloor f\rfloor$,
  which is not defined in case of arithmetic overflow; conversion to and from string values written as
  ``$t$~represents~$i$'', which parses or formats a non-string value and also treating of decimals 
  0.0 and 1.0 as Booleans.

@@

\paragraph{Example 2.} The second example illustrates the remaining parts of the type provision,
including collections and union types. Reusing the \ident{Person} type from the previous example,
consider a collection $[\ident{Person}+\ident{string}]$, which contains a mix of \ident{Person} and string values.

@@

\subsection{Inferring types from values}
\label{sec:formal-inferval}

@@

The key question is, what is a representative sample? Given a set of sample documents, 
the provided type is guaranteed to work if the inferred type of the input is a subtype of any of the
sample documents. Going back to Section~\ref{sec:inference-subtyping}, this means that:
%
\begin{itemize}[noitemsep]
\item[--] Input can contain smaller numerical values (for example, if a sample contains float, the input can contain an integer).
\item[--] Records in the actual input can have additional fields
\item[--] Records in the actual input can have fewer fields, provided that the type of the fields is marked as optional in the sample
\item[--] Union types in the input can have both fewer or more cases
\item[--] When we have a union type in the sample, the actual input can also contain just values of one of the union cases
\end{itemize}
%
The following lemma states that the provided code (generated in Figure~\ref{fig:tp-generation})
works correctly on inputs $s'$ that is a subtype of the sample $s$. More formally, we require that the 
provided expression (using $s'$ as the input) can be reduced to a value and, if it is a class,
all its members can also be reduced to values.

@@

\noindent
Now that we know that the provided types are correct with respect to the subtyping relation, we can
look at the main theorem of the paper. It states that, for any input (which is a subtype of any of the
samples) and any expression $e$, a well-typed program that uses the provided types does not ``go wrong''.

@@

The Lemma~\ref{thm:tp-correctness} guarantees that this does not happen. In Theorem~\ref{thm:safety},
we carefully state that we only consider expressions $e_c$ which ``[do] not contain any structural 
values as sub-expressions''. This makes sure that the only code working with structural values is the
code generated by the type provider.

@@

The \emph{relativized safety} property does not guarantee the same amount of safety as standard type safety
for programming languages without type providers. However, it reflects the reality of programming with external
data sources that is increasingly important in the age of web \cite{age-of-web}. So, type providers do not
reduce the safety -- they simply make the existing issues visible.

@@

Finally, the formal model presented here ignores two aspects of the real type provider mechanisms. As discussed
here, we do not use the \emph{type erasure} mechanism, but treat type providers as if they were actually 
generating code. For XML, CSV and JSON, both models can be used. However, one interesting aspect of erasing 
type providers is that the types can be generated lazily -- only classes that are actually used in the type
checked code are generated. Extending our formalism to capture this aspect is an interesting future work
relevant to other type providers -- including the World Bank type provider that is also available
in the F\# Data library.

@@

\subsection{Parsing structured values}
\label{sec:impl-parsing}

@@

The type provider mechanism has been described before, but this paper is novel in that it explores concrete
type providers from the perspective of programming language theory. This is particularly important as the
F\# Data library is becoming de-facto standard tool for data access in F\#\footnote{At the time of writing, 
the library has 37,000 downloads on NuGet; over 1600 commits and 36 contributors on GitHub.} and other 
languages are beginning to adopt mechanisms similar to F\# type providers.
